Coupling finite and boundary element methods to solve the Poisson--Boltzmann equation for electrostatics in molecular solvation

### 2023-09-11 conclusion.tex
edits to new results

This paper presents the first implementation of a FEM-BEM coupling approach to solve the Poisson--Boltzmann equation for molecular electrostatics. This brings the best of both worlds: the accuracy and efficiency of BEM to exactly enforce the boundary conditions at infinity, and the flexibility of FEM to account for space variations of the material properties and nonlinearities. After presenting verification results for a sphere and arginine with a constant permittivity inside the solute, we showcased our implementation with an advanced modeling technique that considers Gaussian-varying permittivities in a confined region, with the results validated against the widely-used APBS software. The final scaling results for larger molecules start from protein G B1 (955 atoms) and go up to immunoglobulin G (20\,148 atoms), proving the applicability of this approach for realistic problems. Even though our implementation was able to reach medium-to-large systems, we recongnize the need for further research towards better preconditioning of the linear system, optimizing the coupling technique, acceleration algorithms, and parallel execution, especially as we look towards much larger solutes, like viruses. We hope this proof-of-concept work will serve as motivation for future model development that considers space-varying permittivities and Debye lengths, and mixed linear-nonlinear techniques, especially for highly-charged systems, like nucleic acids.

>>>>

This paper presents the first implementation of a FEM-BEM coupling approach to solve the Poisson--Boltzmann equation for molecular electrostatics. This brings the best of both worlds: the accuracy and efficiency of BEM to exactly enforce the boundary conditions at infinity, and the flexibility of FEM to account for space variations of the material properties and nonlinearities. After presenting verification results of solvation energy for a sphere and arginine with a constant permittivity inside the solute, and binding energy  for an extensive data set, we showcased our implementation with an advanced modeling technique that considers Gaussian-varying permittivities in a confined region, with the results validated against the widely-used APBS software. The final scaling results for larger molecules start from protein G B1 (955 atoms) and go up to immunoglobulin G (20\,148 atoms), proving the applicability of this approach for realistic problems. Even though our implementation was able to reach medium-to-large systems, we recongnize the need for further research towards better preconditioning of the linear system, optimizing the coupling technique, acceleration algorithms, and parallel execution, especially as we look towards much larger solutes, like viruses. We hope this proof-of-concept work will serve as motivation for future model development that considers space-varying permittivities and Debye lengths, and mixed linear-nonlinear techniques, especially for highly-charged systems, like nucleic acids.

### 2023-09-11 main.tex
edits to new results

\begin{abstract}
The Poisson--Boltzmann equation is widely used to model electrostatics in molecular systems. Available software packages solve it using finite difference, finite element, and boundary element methods, where the latter is attractive due to the accurate representation of the molecular surface and partial charges, and exact enforcement of the boundary conditions at infinity. However, the boundary element method is limited to linear equations and piecewise constant variations of the material properties. In this work, we present a scheme that couples finite and boundary elements for the Poisson--Boltzmann equation, where the finite element method is applied in a confined {\it solute} region, and the boundary element method in the external {\it solvent} region. As a proof-of-concept exercise, we use the simplest methods available: Johnson--N\'ed\'elec coupling with mass matrix and diagonal preconditioning, implemented using the Bempp-cl and FEniCSx libraries via their Python interfaces. We showcase our implementation by computing the polar component of the solvation free energy of a set of molecules using a constant and a Gaussian-varying permittivity. We validate our implementation against the finite difference code APBS (to 0.5\%), and show scaling from protein G B1 (955 atoms) up to immunoglobulin G (20\,148 atoms). For small problems, the coupled method was efficient, outperforming a purely boundary integral approach. For Gaussian-varying permittivities, which are beyond the applicability of boundary elements alone, we were able to run medium to large sized problems on a single workstation. Development of better preconditioning techniques and the use of distributed memory parallelism for larger systems remains an area for future work. We hope this work will serve as inspiration for future developments that consider space-varying field parameters, and mixed linear-nonlinear schemes for molecular electrostatics with implicit solvent models. 
\end{abstract}

>>>>

\begin{abstract}
The Poisson--Boltzmann equation is widely used to model electrostatics in molecular systems. Available software packages solve it using finite difference, finite element, and boundary element methods, where the latter is attractive due to the accurate representation of the molecular surface and partial charges, and exact enforcement of the boundary conditions at infinity. However, the boundary element method is limited to linear equations and piecewise constant variations of the material properties. In this work, we present a scheme that couples finite and boundary elements for the Poisson--Boltzmann equation, where the finite element method is applied in a confined {\it solute} region, and the boundary element method in the external {\it solvent} region. As a proof-of-concept exercise, we use the simplest methods available: Johnson--N\'ed\'elec coupling with mass matrix and diagonal preconditioning, implemented using the Bempp-cl and FEniCSx libraries via their Python interfaces. We showcase our implementation by computing the polar component of the solvation free energy of a set of molecules using a constant and a Gaussian-varying permittivity. As validation, we compare against well-established finite difference solvers for an extensive binding energy data set, and with the finite difference code APBS (to 0.5\%) for Gaussian permittivities. We also show scaling results from protein G B1 (955 atoms) up to immunoglobulin G (20\,148 atoms). For small problems, the coupled method was efficient, outperforming a purely boundary integral approach. For Gaussian-varying permittivities, which are beyond the applicability of boundary elements alone, we were able to run medium to large sized problems on a single workstation. Development of better preconditioning techniques and the use of distributed memory parallelism for larger systems remains an area for future work. We hope this work will serve as inspiration for future developments that consider space-varying field parameters, and mixed linear-nonlinear schemes for molecular electrostatics with implicit solvent models. 
\end{abstract}

### 2023-09-11 results.tex
edits to new results

Folllowing the data set proposed by Fenley and co-workers,\cite{HarrisBoschitcshFenley2013} available in \url{http://www.sb.fsu.edu/~mfenley/convergence/downloads/convergence_pqr_sets.tar.gz}, and later used by Wei {\it et al.}\cite{nguyen2017accurate} to assess the accuracy of different Poisson-Boltzmann solvers, we tested our method by computing binding energies. 
We first performed a mesh refinement study with the structure of 1biv, available in that same data set, to decide on an appropriate mesh density.
Table XX shows that finite element grids generated from a surface mesh with 12, 17 and 24 vertices per \AA$^2$, and a quality parameter of 1.1 in TetGen, are within the convergence region ({\it ie.} the observed order of convergence agrees with the theoretical order of the method).\cite{roache1998verification,CooperBardhanBarba2014}
From this result, we performed Richardson extrapolation to get an approximate true solution, and found that the mesh with 24 vertices per \AA$^2$ was XX\% away, which we chose for the remaining of this test.

@@

Figures \ref{fig:dna_drug}, \ref{fig:barnase_barstar}, and \ref{fig:rna_peptide} show the binding energies of structures of a DNA-drug complex, the barnase-barstar complex, and a RNA-peptide complex, respectively, presented in Fenley's data set.\cite{HarrisBoschitcshFenley2013} The data for CPB\cite{HarrisBoschitcshFenley2013} and MIBPB, Delphi, APBS and PBSA where taken from the literature,\cite{nguyen2017accurate} for the finest mesh reported. 
From Figures \ref{fig:dna_drug}, \ref{fig:barnase_barstar}, and \ref{fig:rna_peptide} it is evident that our FEM-BEM approach gives results that are within the range of the other software.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{binding_energy_dna-drug.pdf}
\caption{Binding energies for structures corresponding to a DNA-drug complex}
\label{fig:dna_drug}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{binding_energy_barnase-barstar.pdf}
\caption{Binding energies for structures corresponding to the barnase-barstar complex}
\label{fig:barnase_barstar}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{binding_energy_rna-peptide.pdf}
\caption{Binding energies for structures corresponding to a RNA-peptide complex}
\label{fig:rna_peptide}
\end{figure}

>>>>

Folllowing the data set proposed by Fenley and co-workers,\cite{HarrisBoschitcshFenley2013} available in \url{http://www.sb.fsu.edu/~mfenley/convergence/downloads/convergence_pqr_sets.tar.gz}, and later used by Wei {\it et al.}\cite{nguyen2017accurate} to assess the accuracy of different Poisson-Boltzmann solvers, we tested our method by computing binding energies. 
This data set consists of three family of structures: DNA-drug complexes, different versions of the barnase-barstar complex, and RNA-peptide complexes.

According to the arginine results in Figure \ref{fig:arg_constant_energy}, a mesh density of 24 vertices per \AA$^2$ gives a solvation energy that is close to the true solution for a realistic geometry. 
In that case, the difference in solvation energy between calculations using finite element grids generated from a surface mesh of 8.6 and 24.5 vertices per \AA$^2$ (a 3$\times$ refinement) is less than 1\%.
Then, we used surface meshes with 24 vertices per \AA$^2$ and a quality parameter of 1.1 in TetGen to generate the meshes used to compute binding energies in Fenley's data base.

@@

Figures \ref{fig:dna_drug}, \ref{fig:barnase_barstar}, and \ref{fig:rna_peptide} show the binding energies of structures of DNA-drug complexes, barnase-barstar complexes, and RNA-peptide complexes, respectively, presented in Fenley's data set.\cite{HarrisBoschitcshFenley2013} The data for CPB\cite{HarrisBoschitcshFenley2013} and MIBPB, Delphi, APBS and PBSA where taken from the literature,\cite{nguyen2017accurate} for the finest mesh reported. 
From Figures \ref{fig:dna_drug}, \ref{fig:barnase_barstar}, and \ref{fig:rna_peptide} it is evident that our FEM-BEM approach gives results that are within the range of the other software.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{binding_energy_dna-drug.pdf}
\caption{Binding energies for structures corresponding to DNA-drug complexes.}
\label{fig:dna_drug}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{binding_energy_barnase-barstar.pdf}
\caption{Binding energies for structures corresponding to the barnase-barstar complex.}
\label{fig:barnase_barstar}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{binding_energy_rna-peptide.pdf}
\caption{Binding energies for structures corresponding to RNA-peptide complexes.}
\label{fig:rna_peptide}
\end{figure}
